Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-08 paper/metamonads.tex
minor changes to abstract. changes for spacing.

@@

@@

\begin{abstract}
We use monads and Ocaml's advanced module system to
demonstrate that, in a generative context, it is
possible to totally eliminate the overhead of abstraction.
This lets one use extreme forms of \textit{information hiding}
at no run-time cost.  Furthermore the typed nature of
the generative context provide static guarantees about the
generated code.  The various extensions
(aspects) can be made completely orthogonal and compositional,
even in the presence of name-generation for temporaries and
other bindings and ``interleaving'' of aspects.  We also
show how to encode some domain-specific knowledge so that
``clearly wrong'' compositions can be statically rejected by the
compiler.
\end{abstract}

@@

In high-performance and numeric computing, there is a well-known
issue of balancing between maximal performance and the level of
abstraction at which code is written.  Furthermore, already in
linear algebra, there is a wealth of different concerns that 
\emph{may} need to be addressed.  While it is possible to manually
inline and simplify such code, this is labour-intensive and 
error-prone.  Furthermore the resulting code is frequently unreadable,
and thus unmaintainable.

@@

Whether one prefers to code in an imperative, object-oriented, functional, 
or even in a mixed style, all abstraction mechanisms in current
languages have a non-zero run-time cost.  Whether it is procedure,
method or function call, these have a cost; worse still, higher level
and/or higher order abstractions (like Functors or AspectJ's aspects)
frequently have a larger cost still.  Of course, if one is faced with
writing 60 different implementations of the ``same'' algorithm
\oleg{citation?}, this
abstraction cost starts to look cheap!  Although bad programmers have
been known to resort to cut\&paste programming, no self-respecting
programmer would be caught doing that\footnote{Sadly, there are still
many programmers who do not see the harm of cut\&paste programming}.
The common solution is to turn to code generation.

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but it
does not scale well, it is a debugging nightmare, and usually suffers
most from Software Aging \cite{parnas_aging}.  As with other tasks in
programming, the answer is to turn to statically typed languages,
which provide strong guarantees about the soundness of the resulting
code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
out: not only it guarantees well-formedness and the absense of problems 
such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
statically ensures that both the code generator and the generated code
are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
programming in C++ and Template Haskell do offer some amount of static
typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and
  MetaML.} offers stronger guarantees.  As well, since MetaOCaml is
implemented on top of the very solid Ocaml language, one accrues
additional benefits.

@@

\begin{figure*}
\begin{code}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) =
   struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    type v = Dom.v
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = mdo \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        mdo \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = mdo \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( mdo \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} 
    in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code}
\end{figure*}

@@

\oleg{later on, we can write the following just let fpc = .<...>.}

>>>>

\usepackage[medium,compact]{titlesec}

@@

\parskip 0pt plus 1pt minus 2pt
\textfloatsep 4pt plus 2pt minus 4pt % Less space around figures
\abovecaptionskip 0pt plus 0pt minus 2pt
\belowcaptionskip 0pt plus 0pt minus 2pt

\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}   
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

\begin{abstract}
With Gaussian Elimination as a representative family of numerical
and symbolic
algorithms, we use multi-stage programming, monads and Ocaml's
advanced module system to demonstrate the complete elimination of the
abstraction overhead while avoiding any inspection of the generated
code.  We parameterize our Gaussian Elimination code to a great extent
(over domain, matrix representations, determinant tracking, 
pivoting policies, result types, etc) at no run-time cost.  Because
the resulting code is generated just right and not changed afterwards,
we enjoy MetaOCaml's guaranty that the generated code is well-typed.
We further demonstrate that various abstraction parameters (aspects)
can be made orthogonal and compositional, even in the presence of
name-generation for temporaries and other bindings and
``interleaving'' of aspects.  We also show how to encode some
domain-specific knowledge so that ``clearly wrong'' compositions can
be statically rejected by the compiler when processing the generator
rather than the generated code.
\end{abstract}

@@

In high-performance, symbolic, and numeric computing, there is a well-known
issue of balancing between maximal performance and the level of
abstraction at which code is written.  Furthermore, already in
linear algebra, there is a wealth of different concerns that 
\emph{may} need to be addressed.  While it is possible to manually
inline and simplify such code, this is labour-intensive and 
error-prone.  Furthermore the resulting code is frequently unreadable,
and thus unmaintainable.

@@

Whether one prefers to code in an imperative, object-oriented, functional, 
or even in a mixed style, all abstraction mechanisms in current
languages have a non-zero run-time cost.  Whether it is procedure,
method or function call, these have a cost; worse still, higher level
and/or higher order abstractions (like Functors or AspectJ's aspects)
frequently have a larger cost still.  Of course, if one is faced with
writing 60 different implementations of the ``same'' algorithm
\oleg{citation?}, this
abstraction cost starts to look cheap!  Although bad programmers have
been known to resort to cut\&paste programming, no self-respecting
programmer would be caught doing that.
The common solution is to turn to code generation.

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but it
does not scale well, it is a debugging nightmare, and usually suffers
most from Software Aging \cite{parnas_aging}.  As with other tasks in
programming, the answer is to turn to statically typed languages,
which provide strong guarantees about the soundness of the resulting
code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
out: not only it guarantees well-formedness and the absense of problems 
such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
statically ensures that both the code generator and the generated code
are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
programming in C++ and Template Haskell do offer some amount of static
typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml
offers stronger guarantees.  As well, since MetaOCaml is
implemented on top of the very solid Ocaml language, one accrues
additional benefits.

@@

\begin{figure*}
\begin{code}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) = struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = mdo \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        mdo \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = mdo \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( mdo \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code}
\end{figure*}

@@
